---
title: "Getting started with uwb"
vignette: >
  %\VignetteIndexEntry{Getting started with uwb}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    page-layout: full
    mainfont: Open Sans
toc: true
editor: source
---

```{r include=FALSE}
rm(list = ls())
library(tidyverse)
library(uwb)
```

# Introduction

The `uwb` package provides a set of objects, functions, themes, and color palettes for easy and consistent data visualization at the Quality Department of the University of West Bohemia. The main components are:

-   **Objects** — built-in dataset, faculty names and colors, visual settings
-   **Data preparation functions** — `prep_*()` and `mean_bat*()` families for reshaping survey data into a plot-ready format
-   **Plot functions** — `plot_*()` family for bar charts, lollipop plots, stacked and dodge plots, trend plots
-   **Theme** — `theme_uwb()` and `theme_uwb_horiz()` for a consistent visual style
-   **Color scales** — `scale_fill_uwb()` and `scale_color_uwb()` with built-in UWB color palettes

Dedicated vignettes cover themes and colors (*UWB theme and colors*) and data preparation and plot functions (*Data preparation and plotting*) in detail. This vignette focuses on installation, built-in objects, and the general workflow.

# Installation

``` r
# Install remotes if you don't have it yet
install.packages("remotes")

# Install galetap/uwb from GitHub
remotes::install_github("galetap/uwb")

# Reinstall when the package is updated on GitHub
remotes::install_github("galetap/uwb", force = TRUE)
```

After installation, load the package in every session:

``` r
library(uwb)
```

# Objects

The package provides several built-in objects. After loading the package with `library(uwb)`, they are available directly in the session without any further loading step.

## Example dataset: `data`

`data` is a data frame intended for demonstrating the data preparation and plotting functions. Rows are respondents; columns are variables of different types that cover the main use cases of the package:

-   `fak` — faculty (categorical, single choice)
-   `rok` — academic year (categorical, single choice)
-   `pohlavi` — gender (categorical, single choice)
-   `typ` — type of study (categorical, single choice)
-   `num1`–`num5` — numeric ratings (e.g. satisfaction on a 0–10 scale)
-   `mc1`–`mc5` — multiple choice items (0/1 coded: 1 = chosen)
-   `bat1`–`bat5` — battery items with identical ordered response options (e.g. agree / neutral / disagree)

```{r}
head(data)
```

```{r}
glimpse(data)
```

## Faculty and institute names: `faks` and `parts`

`faks` is a named character vector mapping short faculty codes to their full Czech names. The names of the vector are the short codes used in data (e.g. `"FAV"`); the values are the full names for plot labels.

```{r}
faks
```

`parts` is the equivalent for the non-faculty university institutes (e.g. NTC, UJP).

```{r}
parts
```

These vectors are useful for recoding factor levels in raw data to display-ready labels, and they define the order of faculties in faculty-scale plots.

## Visual settings: `uwb_vals`

`uwb_vals` is a list (implemented internally as an environment) that stores the default values controlling the visual appearance of themes and plots — font family, text sizes, line widths, point sizes, and fixed colors used outside aesthetic mappings. The `plot_*()` functions and `theme_uwb()` read these values internally, so changing a value in `uwb_vals` affects all subsequent plots without any additional arguments.

```{r}
as.list(uwb_vals)
```

Example: increase the base text size globally.

``` r
uwb_vals$tsize <- 12
```

## University color: `c_zcu`

`c_zcu` is the official dark-blue color of the University of West Bohemia as a hex string. It is used as the default single-color in plots.

```{r}
c_zcu
```

## Faculty and institute colors: `c_faks` and `c_parts`

`c_faks` is a named character vector with the official colors of the nine faculties, in the same alphabetical order as `faks`. `c_parts` is the equivalent for non-faculty institutes.

```{r}
c_faks
```

```{r}
c_parts
```

These vectors can be used directly when a manual color assignment is needed — for example, to highlight a single faculty in a plot using `scale_fill_manual(values = c_faks)`.

## Color scale definitions: `uwb_scales`

`uwb_scales` is a named list of color hex vectors that define the built-in color scales. It is used internally by `scale_fill_uwb()` and `scale_color_uwb()`, but can also be inspected or used directly to build custom color sequences.

```{r}
names(uwb_scales)
```

The available scales are: `quali`, `faks`, `uwb_faks`, `mono`, `mono_rev`, `monored`, `monored_rev`, `long`, `long_rev`, `bi`, `bi_rev`. See the *UWB theme and colors* vignette for visual examples of all scales.

# Overview of functions

## Data preparation functions

The `prep_*()` functions transform a raw data frame into a plot-ready tibble. Each function computes relative frequencies (percentages), generates text labels and their positions, and automatically imports plot titles from a `codebook` object if one exists in the environment (see [The codebook object] below).

| Function | What it does |
|---|---|
| `prep_single()` | Distribution of a single categorical variable |
| `prep_gr()` | Distribution of a categorical variable by one grouping variable |
| `prep_gr2()` | Distribution of a categorical variable by two grouping variables |
| `prep_mc()` | Distribution of a multiple choice question (0/1 coded columns) |
| `prep_mc_gr()` | Multiple choice distribution by a grouping variable |
| `prep_bat()` | Distribution of a battery of items with identical response options |
| `prep_bat_gr()` | Battery distribution by a grouping variable |
| `mean_bat()` | Mean scores of a set of numeric variables |
| `mean_bat_gr()` | Mean scores by one grouping variable |
| `mean_bat_gr2()` | Mean scores by two grouping variables |

### Auxiliary data preparation functions

These functions are called internally by the `prep_*()` functions but can also be used directly for custom workflows.

| Function | What it does |
|---|---|
| `generate_textlabs()` | Generates text label columns (`labvar`, `pos`, `cvar_text`) from `yvar` and `n` |
| `impute_labs()` | Imports `title`, `subtitle`, `caption`, and `lab` from a `codebook` object |
| `polish_var()` | Wraps long labels and drops unused factor levels in a grouping variable |

## Plot functions

| Function | What it does |
|---|---|
| `plot_bar()` | Bar chart (horizontal by default) |
| `plot_lolli()` | Lollipop chart |
| `plot_stack()` | Stacked bar chart for grouped data |
| `plot_dodge()` | Dodge (side-by-side) bar chart for grouped data |
| `plot_trend()` | Line + point trend plot for a single variable over time |
| `plot_trend_gr()` | Trend plot for multiple variables simultaneously |
| `plot_out()` | Bar chart highlighting faculties with extreme values |
| `plot_ribbon()` | Ribbon plot |

## Exploratory functions

`ex1()` and `ex2()` are quick helpers for inspecting the distribution of one or two variables in a data frame during data exploration (counts and percentages printed to the console).

```{r}
ex1(data, fak)
```

```{r}
ex2(data, fak, typ)
```

# General workflow

The typical workflow for producing a plot has two steps: (1) prepare the plot data with a `prep_*()` function, then (2) pass it to a `plot_*()` function.

```{r echo=TRUE, warning=FALSE}
# Step 1: prepare the plot data
plot_data <- prep_single(data, fak)
head(plot_data)
```

```{r echo=TRUE, warning=FALSE}
# Step 2: plot
plot_bar(plot_data)
```

The two steps can be combined in a single pipeline. The pipeline also makes it easy to add or modify columns between preparation and plotting — for example, to insert a plot title manually:

```{r echo=TRUE, warning=FALSE}
prep_single(data, fak) |>
  mutate(title = "Distribution by faculty") |>
  plot_bar()
```

By default, `plot_bar()` produces a horizontal bar chart. Set `horiz = FALSE` for a vertical chart:

```{r echo=TRUE, warning=FALSE}
prep_single(data, fak) |>
  mutate(title = "Distribution by faculty") |>
  plot_bar(horiz = FALSE)
```

## The codebook object

Each `prep_*()` function looks for an object named `codebook` in the calling environment and tries to import plot labels (`title`, `subtitle`, `caption`, and `lab` for batteries) from it. If no such object exists, those columns are filled with empty strings.

The `codebook` must be a data frame with the following columns:

-   `name` — variable name as it appears in the dataset; used to match rows to the plot data
-   `label` — full variable label or question wording
-   `lab` — short item label for battery and multiple choice questions; `NA` for other variables
-   `title` — plot title text
-   `subtitle` — sample description (e.g. "UWB graduates, N = 450")
-   `caption` — footnote text

The package ships with a built-in `codebook` object that provides labels for the `data` dataset:

```{r}
print(codebook)
```

Once `codebook` exists in the environment, the `prep_*()` functions pick it up automatically:

```{r echo=TRUE, warning=FALSE}
prep_single(data, fak) |>
  plot_bar()