---
title: "Data preparation and plotting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Data preparation and plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width  = 7,       # inches — matches the html_vignette content width
  fig.height = 4.3,     # shorter, wider aspect ratio (~16:10)
  fig.retina = 2,       # crisp rendering on HiDPI screens
  dpi        = 96,      # web-appropriate resolution
  out.width  = "100%"   # scale the figure to exactly the text-column width
)

```

```{r include=FALSE}
library(tidyverse)
library(uwb)
```

# Introduction

Producing a chart with the `uwb` package is a **two-step workflow**: first you transform your raw, respondent-level data into a plot-ready tibble with a `prep_*()` or `mean_*()` function, then you pass that tibble to a matching `plot_*()` function. Because the preparation functions always return the same standardised structure (a `yvar` with the value to be plotted, an `xvar` with the category, label columns, and so on), the plot functions know exactly how to consume it, and the two steps chain together naturally in a single pipeline.

This vignette walks through both steps together. For each type of survey variable it shows the appropriate preparation function and then the plot it is designed for, so you can see the full path from data to finished chart.

It covers the full family of preparation functions:

-   `prep_single()`, `prep_gr()`, `prep_gr2()` — distributions of a categorical variable, optionally split by one or two grouping variables.
-   `prep_mc()`, `prep_mc_gr()` — distributions of multiple choice questions (binary indicators of whether the item was chosen).
-   `prep_bat()`, `prep_bat_gr()` — distributions of batteries of items that share the same response options.
-   `mean_bat()`, `mean_bat_gr()`, `mean_bat_gr2()` — mean scores of numeric items, optionally split by one or two grouping variables.

Each section pairs these with the relevant plotting functions — `plot_bar()`, `plot_lolli()`, `plot_stack()`, and `plot_dodge()` — and the vignette closes with a summary table mapping every preparation function to its typical plot. 

The three auxiliary helpers (`generate_textlabs()`, `impute_labs()`, `polish_var()`) that the preparation functions call internally are also described. Arguments that feed these helper functions are available in all prep_functions. They are not described repeatedly in each section but in the [auxiliary functions](#auxiliary-functions) section.

All functions have data (`dat`) as the first argument. It is not repeated in the instructions below.

All examples use the built-in `example_data` dataset and the built-in `codebook` (see the *Getting started with uwb* vignette for their structure, and *UWB theme and colors* for styling the resulting plots).

# The structure of a prepared tibble

Before going through each function, it helps to understand the columns that every preparation function produces. The exact set differs slightly between functions, but the most important ones are here:

| Column | Meaning |
|---|---|
| `yvar` | The numeric value to be plotted — a **percentage** for the `prep_*()` functions, a **mean** for the `mean_*()` functions |
| `n` | The raw count behind each `yvar` (frequency tables only) |
| `nsize` | The group size (the denominator of the percentage / number of valid answers) |
| `xvar` | The category placed on the category axis (polished version with wrapped lines and sample size info) |
| `zvar` | A secondary category used for stacking / grouping (polished version) |
| `zzvar` | An additional grouping category (polished version), used mostly for facetting |
| `xvar_df`, `zvar_df`, `zzvar_df`| Crude (default) versions of the variables before polishing |
| `name` | The original variable name; used to match labels from the `codebook` |
| `labvar`, `labvar_single`, `pos`, `pos_single`, `cvar_text` | Text-label content and positions, added by `generate_textlabs()` |
| `title`, `subtitle`, `caption`, `lab` | Plot labels, added by `impute_labs()` from the `codebook` |

Because the output is just a tibble, you can always inspect it, add columns, or override labels with `mutate()` before piping it into a `plot_*()` function.

# Distribution of a categorical variable

## `prep_single()` — a single variable

`prep_single()` computes the distribution (relative frequencies, in percent) of one categorical variable.

```{r}
prep_single(example_data, fak)
```

Key arguments:

-   `drop_na` — when `TRUE` (the default), missing values of the variable are dropped before computing percentages.
-   `order` — when `TRUE`, the categories are reordered by frequency (largest first), which is convenient for ranked bar or lollipop charts.
-   `show_n` — when `TRUE` (the default), sample size ("N=") is show in the subtitle.

## `plot_bar()`, `plot_lolli()` - plotting distributions of a single categorical variable

The typical use of `prep_single` is to pipe the result straight into `plot_bar()` or `plot_lolli()`:

```{r warning=FALSE}
prep_single(example_data, fak) |>
  plot_bar()
```

```{r warning=FALSE}
prep_single(example_data, fak, order = TRUE) |>
  plot_lolli()
```


Key argument:

-   `horiz` — when `TRUE` (the default), the categories are plotted on the vertical axis. 

Horizontal plots are especially useful if the number of 'xvar' values is large or if the names into onto several lines. It is easier to expand the height of the plot than its width. Vertical plot is fine when the labels are short.

```{r echo=TRUE, warning=FALSE}
prep_single(example_data, fak, order = TRUE) |>
  plot_bar(horiz = FALSE)
```

## `prep_gr()`, `prep_gr2()` — split by one or two grouping variables

`prep_gr()` computes the distribution of `var` within each level of a grouping variable `grvar`. Percentages always sum to 100 % inside each group.

```{r}
prep_gr(example_data, typ, fak)
```

Key arguments:

-   `add_total` — if `TRUE` (the default) an overall group is added. 
-   `lab_total` — label to be used for the total. Default is `ZČU`.

```{r}
prep_gr(example_data, typ, fak, add_total = TRUE) |> pull(xvar_df) |> unique()
```

`prep_gr2()` extends `prep_gr()` with a second grouping variable `grvar2`. The first grouping variable (`grvar`) becomes the category axis, `var` becomes the stacking/dodging variable, and `grvar2` is kept as `zzvar` for use as a facet. It accepts the same arguments as `prep_gr()`.

The grouped output is meant for `plot_stack()` or `plot_dodge()`.

## `plot_stack()` - plot with stacked bars

```{r warning=FALSE}
prep_gr(example_data, pohlavi, fak, add_total = TRUE) |>
  plot_stack()
```

Key arguments:

-   `horiz` — if `TRUE` (the default) categories are plotted on the vertical axis.
-   `lab_min` — Minimum slice size (percentage) below which the label is suppressed (set to `NA`). Default `.uwb_vals$lim_stack_no`.

Customizing is the `lab_min` is useful in facet plots or when the category labels are long and the stacked bars get squeezed more than usual.

```{r}
prep_gr2(example_data, pohlavi, fak, typ) |> 
  filter(!is.na(pohlavi)) |> 
  plot_stack(lab_min = 10) + facet_wrap(~ zzvar, scales = "free_y")
```

## `plot_dodge()` - plot with dodged parts

```{r warning=FALSE}
prep_gr(example_data, pohlavi, fak) |>
  plot_dodge()
```

Key argument:

-   `horiz` — if `TRUE` (the default) categories are plotted on the vertical axis.

# Multiple choice questions

Multiple choice items are stored as a set of 0/1 (or other binary coded) columns, one per option . The `prep_mc*()` functions summarize how often each option was chosen.

## `prep_mc()` — overall multiple choice distribution

`prep_mc()` takes the columns of the multiple choice block in `vars` and reports, for each item, the share of respondents who selected the value. Each row is one item; `yvar` is the percentage who chose it. When a `codebook` is available, the item labels (`lab`) are used as the category labels.

```{r}
prep_mc(example_data, vars = c("mc1", "mc2", "mc3", "mc4", "mc5"), chosen = 1)
```

Key arguments:

-   `vars` — the columns that represent the possible choices (items) (defaults to all columns of `dat`).
-   `chosen` — the value that counts as "selected".
-   `order` — reorder items by frequency.

Plots for single distribution are suitable for plotting the outcome:

```{r warning=FALSE}
prep_mc(example_data, vars = c("mc1", "mc2", "mc3", "mc4", "mc5"),
        chosen = 1) |>
  plot_bar()
```

```{r warning=FALSE}
prep_mc(example_data, vars = c("mc1", "mc2", "mc3", "mc4", "mc5"),
        chosen = 1, order = TRUE) |>
  plot_lolli()
```

## `prep_mc_gr()` — multiple choice by a grouping variable

`prep_mc_gr()` computes the same selection percentages within each level of a grouping variable. The grouping variable goes on the category axis; the items become the secondary category (`zvar`).

```{r}
prep_mc_gr(example_data, vars = c("mc1", "mc2", "mc3", "mc4", "mc5"), grvar = fak, chosen = 1)
```

Key arguments:

-   `vars`, `chosen`, `drop_na` — same as in `prep_mc`.
-   `add_total` — if `TRUE` (the default) an overall group is added. 
-   `lab_total` — label to be used for the total. Default is `ZČU`.

It is usually paired with dodge plot or facetting to compare how often each item was chosen across subgroups. The `nsize` is shown for the groups (the `zvar`).

```{r}
# Dodge plot fits without changes
prep_mc_gr(example_data, vars = c("mc1", "mc2", "mc3", "mc4", "mc5"), grvar = fak, chosen = 1) |> 
  plot_dodge() 

# Facet plot needs renaming the variables to put grouping var (xvar) in the facets
prep_mc_gr(example_data, vars = c("mc1", "mc2", "mc3", "mc4", "mc5"), grvar = fak, chosen = 1) |> 
  rename(xvar = zvar, zvar = xvar) |> 
  plot_bar() + facet_wrap(~ zvar)
```

# Batteries of categorical items

A battery is a block of variables that share the same response scale, e.g. a set of agree/neutral/disagree statements. The `prep_bat*()` functions report the distribution of the response options for every item.

## `prep_bat()` — distribution of a battery

`prep_bat()` pivots the battery columns to long form and computes, for each item, the percentage falling into each response category. The item name becomes `xvar` and the response value becomes `zvar`.

```{r}
prep_bat(example_data, vars = c("bat1", "bat2", "bat3", "bat4", "bat5"))
```

Key arguments:

-   `show_n` — when `TRUE` (the default), sample size ("N=") is show in the subtitle.
-   `vars` — the battery item columns.
-   `drop_na` — drop missing answers before computing percentages.

This output is designed for `plot_stack()`, which produces one stacked bar per item:

```{r warning=FALSE}
prep_bat(example_data, vars = c("bat1", "bat2", "bat3", "bat4", "bat5")) |>
  plot_stack()
```

*Note: this is not a suitable color pallette for degree of agreement. The use of alternative pallettes is explained in the *UWB theme and colors vignette.* 

## `prep_bat_gr()` — battery split by a grouping variable

`prep_bat_gr()` adds a grouping variable. Here the response value is `zvar`, the item name is `xvar`, and the grouping variable (`zzvar`) is kept for faceting.

```{r}
prep_bat_gr(example_data,
            vars = c("bat1", "bat2", "bat3", "bat4", "bat5"),
            grvar = pohlavi)
```

Key arguments:

-   `add_total` — if `TRUE` (the default) an overall group is added. 
-   `lab_total` —  label to be used for the total. Default is `ZČU`.

The outcome is suitable for a plotting single items (groups compared to each other and optionally to the total) or a facet plot with all items shown separately for each group.

```{r}
# Default settings showing N for each item
prep_bat_gr(example_data |> drop_na(pohlavi),
            vars = c("bat1", "bat2", "bat3", "bat4", "bat5"),
            grvar = pohlavi) |> 
  plot_stack() + facet_wrap(~ zzvar, scales = "free_y")  

# Custom settings showing N for facets (better in this case because all items have the same N)
prep_bat_gr(example_data |> drop_na(pohlavi),
            vars = c("bat1", "bat2", "bat3", "bat4", "bat5"),
            grvar = pohlavi, 
            show_nsize = FALSE, show_nsize_zzvar = TRUE) |> 
  plot_stack() + facet_wrap(~ zzvar, scales = "free_y")  
```

See [the polish_var section](#polish_var-wrap-labels-and-append-sample-sizes) for instruction ho to move the sample size information to facet titles instead.

# Mean scores

When items are numeric, it is often more informative to report a mean than a full distribution. The `mean_*()` functions compute item means and return them in the same plot-ready shape, with `yvar` holding the mean instead of a percentage.

The mean function parses the item values with `readr::parse_number()`, treating the strings listed in `na_lab` (by default `"NA"` and `"Bez odpovědi"`) as missing. This lets them work both with numeric columns and with labelled/character columns that contain non-numeric "no answer" categories.

## `mean_bat()` — means of a battery

`mean_bat()` returns the mean of each numeric item in `vars`. It works with single item, as well.

```{r}
mean_bat(example_data, vars = c("num1", "num2", "num3", "num4", "num5"))
```

Key arguments:

-   `vars` — the numeric item columns.
-   `round_places` — number of decimals used for the rounded label (`labvar_full`).
-   `na_lab` — character values to treat as missing when parsing numbers. Defaults to `c("NA", "Bez odpovědi")`.
-   `order` — when `TRUE`, the categories are reordered by frequency (largest first).

The result pairs with `plot_lolli()` or `plot_bar()`:

```{r warning=FALSE}
# Default settings
mean_bat(example_data, vars = c("num1", "num2", "num3", "num4", "num5")) |>
  plot_bar()

# Customised
mean_bat(example_data, vars = c("num1", "num2", "num3", "num4", "num5"), 
  order = TRUE, show_n = FALSE, round_places = 1) |>
  plot_bar()
```

## `mean_bat_gr()`, `mean_bat_gr2()` — means by one or two grouping variables

`mean_bat_gr()` computes item means within each level of a grouping variable. The grouping variable becomes the category axis (`xvar`, optionally with `n=` appended), and the item name becomes the secondary category (`zvar`).

```{r}
mean_bat_gr(example_data, vars = c("num1", "num2", "num3", "num4", "num5"), grvar = rok)
```

Key arguments:

-   `vars`, `round_places`, `na_lab` — same as `mean_bat`.
-   `add_total` — if `TRUE` (the default) an overall group is added. 
-   `lab_total` —  label to be used for the total. Default is `ZČU`.

`mean_bat_gr2()` adds a second grouping variable. The first grouping variable (`grvar`) becomes `xvar`, the second (`grvar2`) becomes `zvar`, and item means are computed for every combination.

```{r}
mean_bat_gr2(example_data,
             vars = c("num1", "num2", "num3", "num4", "num5"),
             grvar = fak, grvar2 = pohlavi)
```

`plot trend` or `plot_out` are typical endpoints.

## `plot_trend`, `plot_trend_gr` - plots of numerical values over time

`plot_trend()` draws a line-and-point chart where `xvar` is the time axis and `yvar` is the value — typically mean or percentage — over several time points. It has no additional arguments beyond `dat`.

```{r}
prep_gr(example_data, pohlavi, rok) |> 
  filter(zvar_df == "Muž") |> 
  plot_trend() + 
    ylim(0, 100) + 
    labs(title = "Podíl mužů mezi studujícími")
```

`plot_trend_gr()` extends this to multiple groups: `zvar` defines the groups and each gets its own line, making it straightforward to compare trends across subgroups or across indicators.

```{r}
# Compare indicators
mean_bat_gr(example_data, vars = c("num1", "num2", "num3", "num4", "num5"), grvar = rok, add_total = TRUE) |> 
  plot_trend_gr()

# Compare percentages (type sizes) over time
prep_gr(example_data, typ, rok) |> 
  plot_trend_gr()
```

Key arguments:

-   `total` — optional label of the group (level of `zvar`) to be emphasized with larger points, e.g. the overall total.
-   `add_points` — when `TRUE` (the default), points and value labels are drawn on top of the lines.

## plot_out - bubbles with central tendency and highlighted outliers

`plot_out()` is a bubble plot designed to assess heterogeneity across groups. It plots the overall total value (usually mean or percentage) and group-specific values as smaller points jittered around it so that overlapping points remain visible. Group points are muted (light grey) if they are close to the total value. Outlier groups are highlighted. The default setting reflects the most common use case, i.e. showing the heterogeneity of faculties.

The input data must contain groups (faculties) and a total row — typically produced by `prep_gr2()`, `prep_bat_gr` or `mean_bat_gr()` with `add_total = TRUE`. It  is necessary to fiddle with renaming the varialesa bit before sending them to the `plot_out` function. The group variably (typically faculty) must be stored in the `zvar` and `zvar_df`.

Key arguments:

-   `out` — threshold: groups differing from the total by more than this value are highlighted as outliers.
-   `horiz` — when `TRUE`, the chart is drawn horizontally; default is `FALSE`.
-   `total` — label of the total group (level of `zvar_df`) used as the reference. Default is `ZČU`.
-   `add_line` — when `TRUE` (the default), a line connects the total reference values across categories.
-   `alpha_shadow` — transparency of the non-highlighted (shadow) group points. Default `0.15`.

```{r}
# Compare means of a set of numeric indicators
# show_n = FALSE keeps zvar clean so that the codition zvar == "ZČU" inside the plot_out function matches 
mean_bat_gr(example_data, vars = c("num1", "num2", "num3", "num4", "num5"), grvar = fak,
  add_total  = TRUE, show_n = FALSE) |>          
  rename(xvar = zvar, zvar = xvar, xvar_df = zvar_df, zvar_df = xvar_df # renaming makes it fit the plot function 
  ) |> 
    plot_out(out = 0.5, add_line = FALSE) + ylim(0, 10)

# Gender composition - heterogeneity across faculties and types 
# The desired group var (typically fak) has to be grvar to assing total category correctly as a value of zvar (fak) not zzvar (typ)
prep_gr2(example_data, pohlavi, fak, typ, add_total = TRUE, show_nsize = FALSE) |> 
  filter(zvar_df == "Muž") |> 
  select(-zvar, -zvar_df) |> 
  rename(xvar = zzvar, xvar_df = zzvar_df, zvar_df  = xvar_df, zvar = xvar) |> 
  plot_out(out = 10) + 
    labs(
      title = "% mužů mezi studujícími", 
      y = "%"
    )

# Compare percentages in a battery
prep_bat_gr(example_data, vars = c("bat1", "bat2", "bat3", "bat4", "bat5"), grvar = fak, 
  show_nsize = FALSE, add_total = TRUE) |> 
  filter(zvar_df == "Souhlasím") |> 
  select(-zvar, -zvar_df) |> 
  rename(zvar = zzvar, zvar_df = zzvar_df) |> 
  plot_out(out = 5) + 
  labs(
    title = "% studujících, kteří souhlasí s těmito výroky", 
    y = "%")

```

# Auxiliary functions

The preparation functions share three internal helpers. They are exported, so they can also be used directly in custom workflows.

## `generate_textlabs()` — text labels and positions

`generate_textlabs()` takes a data frame that already contains a `yvar` column and adds the columns used by `geom_text()` in the plot functions:

-   `labvar_full` — the rounded value of yvar.
-   `labvar_single` — a character label for single-series bar charts (values below 0.5 are shown as `"<1"`).
-   `labvar` — a character label used by stacked / dodged charts (very small slices are set to `NA` so they are not drawn).
-   `pos`, `pos_single` — the vertical position of each label.`pos_single` is placed above the bar if yvar is low (the text label would not fit into the bar)
-   `cvar`, `cvar_text` — the color of the label, chosen for contrast against the bar in `plot_bar()`.

Arguments:

-   `dat` — a data frame to be modified. It has to contain the yvar.
-   `round_places` — it controls the number of decimals.

```{r}
tibble(yvar = c(5, 42, 0.3, 88)) |>
  generate_textlabs() |> 
  print()
```

## `impute_labs()` — labels from the codebook

`impute_labs()` looks for a `codebook` object in the environment and joins its `title`, `subtitle`, `caption`, `label`, and `lab` columns onto the data by the `name` column. If no `codebook` is found, those columns are added as empty strings. Every `prep_*()` and `mean_*()` function calls it automatically, which is why prepared data already carry plot titles when a codebook is available.

Arguments:

-   `dat` — a data frame to be modified.

```{r}
tibble(name = c("num1", "pohlavi")) |>
  impute_labs() |> 
  print()
```

## `polish_var()` — wrap labels and append sample sizes

`polish_var()` turns a "raw" category column that appears in the data into the final, plot-ready factor while preserving the original order of the values.

Arguments:

-   `dat` — a data frame (containing an `nsize` column to be used if the `nsize` argument is `TRUE`).
-   `var_origin` — name of the raw category column to polish. Default `"xvar_df"`.
-   `var_final` — name of the resulting plot-ready factor column. Default `"xvar"`.
-   `wrap` — when `TRUE` (the default), long labels are wrapped onto several lines.
-   `chrnum` — the number of characters per line used when wrapping labels. Default uwb_vals$chrnum
-   `nsize` — when `TRUE` (the default), the group size (`n = ...`) is appended to each label. It is colour-coded by reliability (grey for large enough groups, dark red for n < 30, bright red for n < 20).

The `x_wrap` / `x_chrnum` / `x_nsize` (and `z_*` or `zz_*` if applicable) arguments are available in all `prep_*` functions and passed straight to `polish_var()`. The default for `nsize` is `FALSE` in `prep_single` and `prep_mc` because they show the N size in the subtitle. The grouped `prep_*` have `nsize = TRUE` by default. The `zvar` and `zzvar` do not show nsize. If desired, it has to be changes manually after calling the `prep_*` function.

```{r}
# With default arguments
tibble(
  xvar_df = c("A long variable name that des not fit the plot", "A short name"),
  nsize = c(15, 320)
) |>
  polish_var() |> print()

# With custom arguments
tibble(
  somevar_df = c("A long variable name that des not fit the plot", "A short name"),
  nsize = c(15, 320)
) |>
  polish_var(var_origin = "somevar_df", var_final = "somevar", chrnum = 50) |> print()

# Changing the parameter from a prep_function
# show_nsize modifies the nsize parameter
# prep_bat_gr allow adding sample size to zzvar, too (it is the only function that does this)
prep_bat_gr(example_data |> drop_na(pohlavi),
            vars = c("bat1", "bat2", "bat3", "bat4", "bat5"),
            grvar = pohlavi, show_nsize = FALSE, show_nsize_zzvar = TRUE) |> 
  plot_stack() + facet_wrap(~ zzvar, scales = "free_y") 


```

# Summary

| Function | Output value (`yvar`) | Typical plot |
|---|---|---|---|
| `prep_single()` | Percentage of one variable | `plot_bar()`, `plot_lolli()` |
| `prep_gr()` | Percentage within a group | `plot_stack()`, `plot_dodge()` |
| `prep_gr2()` | Percentage within a group |  `plot_stack()`, `plot_dodge()` + facets |
| `prep_mc()` | Percentage choosing each option | `plot_bar()`, `plot_lolli()` |
| `prep_mc_gr()` | Percentage choosing each option | `plot_stack()`, `plot_dodge()` |
| `prep_bat()` | Percentage of each battery item | `plot_stack()` |
| `prep_bat_gr()` | Percentage of each battery item | `plot_stack()` + facets |
| `mean_bat()` | Mean of numeric items | `plot_lolli()`, `plot_bar()` |
| `mean_bat_gr()` | Mean within a group | `plot_dodge()`, `plot_lolli()` |
| `mean_bat_gr2()` | Mean within a group | `plot_dodge()` + facets |

In every case the workflow is the same: choose the preparation function that matches the variable type and the desired grouping, inspect or adjust the resulting tibble if needed, and pipe it into the matching `plot_*()` function. See the *UWB theme and colors* vignette for styling, and *Getting started with uwb* for the overall package overview and the `codebook` workflow.