---
title: "UWB vignette"
format: 
  html:
    page-layout: full
    mainfont: Arial
toc: true
editor: source
---

```{r include=FALSE}
rm(list=ls())
library(tidyverse)
library(colorblindcheck) #palette check for color blindness
library(uwb)
```

# Introduction

The package includes a set of objects, functions, themes, and color palettes for easy and consistent data visualization at the Quality Department of the University of West Bohemia.

# Installation

``` r
# Install remotes if you don't have it yet
install.packages("remotes")

# Install galetap/uwb from GitHub
remotes::install_github("galetap/uwb")

# Reinstall when the package is updated on GitHub
remotes::install_github("galetap/uwb", force = TRUE)

```

# Objects

The package includes a set of usefull objects. They can be accessed by

`example_data`

`faks`, `faks_full`: Named vectors with faculty names. The full version includes the non-faculty institutes.

`c_faks`, `c_faks_full`: Named vectors with faculty colors. The full version includes the non-faculty institutes.

# Data prep functions

`prep_single()` distribution of a discrete variable

`prep_gr()` distribution of a discrete variable by another (grouping) variable

`prep_gr2()` distribution of a discrete variable by two grouping variables

`prep_bat()` distribution of a multiple choice question or a battery of questions with identical options

`prep_bat_gr()` distribution of a multiple choice question or a battery of questions with identical options by a grouping variable

### Plot functions

`plot_bar()`

`plot_lolli()`

### Themes

`theme_uwb()`

`theme_uwb_horiz()`

The `horiz` theme is for plots with flipped axes (e.g. horizontal bar charts).

### Color palettes

The palettes work for discrete data only (i.e. isolated values without decimal places). The number of categories is not limited.

`scale_fill_uwb()`

`scale_color_uwb()` and `scale_colour_uwb()`

`uwb_palettes()`

## The general workflow

There will be the `uwb` package later. But for now, we have to source its functions manually. We need 4 scripts in the code folder: set_all.R, set_theme.R, set_colors.R, set_functions.R. The set_all.R script sets some shared values (incl. font and sizes of text elements in theme and plots) and sources the remaining 3 scripts. The font and sizes of text elements may be changed in the set_all.R script before sourcing it.

The general workflow of an analysis is this.

-   Load the `uwb` package/source`set_all.R` and a clean dataset(s)

```{r echo = TRUE, warning = FALSE, output = FALSE}
# Loading the library - WHEN IT IS DONE
# library(uwb) 

# Alternative to the library
 source("code/set_all.R") 
```

-   Load the data (rows = respondents or other units of data collection)

```{r echo = TRUE, warning = FALSE, output = TRUE}
 load("data/example_data.RData")
 head(example_data)
 # summary(example_data)
```

-   Generate the desired plots in two steps:
    -   prepare the plot data with a function; customise data if necessary

    -   plot the data

```{r echo=TRUE, warning=FALSE, results='hide'}
# Bar plot of the variable 'question'
plot_data = prep_single(example_data, question) 
head(plot_data)
plot_bar(plot_data)
```

Note that the plotted variable includes NAs. They are removed from the plot by default. This can be changed in options - see \[Options of the prep\_\*() functions\]. The data generated by `prep_single` adn any other `prep_*()` function include columns for plot labs (title, subtititle and caption) but they are empty because there is no object named 'codebook' to import the labs from - see [Title, subtitle, and caption].

The plot data generation and plotting can be done within one pipeline. No customization of the data is necessary in most cases. The customization is needed for instance when inserting the plot title manually.

```{r echo=TRUE, warning=FALSE, results='hide'}
prep_single(example_data, question)  %>% 
  mutate(title = "Answer to a question") %>% 
  plot_bar()
```

The preparation of plot data and generating plots can be effectively done within loops when we need a set of plots of the same type.

## Structure of the plot data generated by the `prep_*()` functions

All `prep_*()`functions include the following variables.

-   `xvar`, `yvar`: Coordinates of the data points to be plotted. `yvar` gives the relative size (percent) of the respective category.

-   `n`: Absolute size of the category.

-   `labvar_single` and/or `labvar`: Text labels for data points. Some prep functions produce both of them to be compatible with more plot functions.

-   `pos_single` and/or `pos`: Position of text labels. Some prep functions produce both of them to be compatible with more plot functions.

-   `nsize`: Total of which the percents are computed.

-   `nsize_raw`: Number of rows of the data frame that entered the function. It is the same as `nsize` if option drop_na == FALSE (the default is TRUE).

-   `title`, `subtitle`, `caption`: Labs for plot.

Some functions generate further variables.

-   `zvar`: Third varible to define main aesthetics (usually fill)

-   `cvar`: Fill color outside the aesthetics (in bar plots or lollipop plots)

-   `cvar_text`: Color of text labels

## Title, subtitle, and caption

Each `data_prep*()` function looks for an object called `codebook` and tries to import lab data from it. If no such object exists, they are filled with empty quotes. In any case, the values are constant across rows. The `plot_*()` functions look in the first row for the labs.

The `codebook` object is a data frame with following columns:

-   `name`: Variable name as it appears in the dataset (small case letters, no spaces). This column is used to merge the codebook to the plot data.

-   `title`, `subtitle`, `caption`: Text that should be imported to the plot. The title usually includes the wording of the question; the subtitle information about the sample (whole sample vs. subsample and N).

-   `lab`: only for batteries and multiple choice questions, otherwise NA. It gives the item label (the wording of the item).

For an example see [Lollipop plot of a multiple choice question, labs imported from codebook].

## Options of the `prep_*()` functions

`drop_na` If TRUE, it drops the NAs. Default is TRUE.

`order` If TRUE, it orders categories from the most to the least frequent. Default is FALSE

`add_total` For grouped data only. If true, it adds total as the first category. Default is FALSE.

## Option of the `plot_*()` functions {data-link="Options of the prep_*() functions"}

`horiz` if TRUE, the plot has flipped axes, i.e. the y-axis is horizontal. Default is TRUE.

```{r echo=TRUE, warning=FALSE, results='hide'}
prep_single(example_data, var = question, drop_na = FALSE, order = TRUE)  %>% 
  mutate(title = "Answer to a question",
         ) %>% 
  plot_bar(horiz = FALSE)
```

## Color and Fill scales

There are 5 built-in color scales that can be called from ggplot by `scale_fill_uwb("scale_name")` or `scale_color_uwb("scale_name")`.

```{r include=FALSE}

#fake data to plot scales
cd=tibble(yvar=rep(1,10),
          xvar=letters[1:10])

#starting plots to be built upon
cplot=ggplot(cd,aes(xvar,yvar)) +
  geom_col(aes(fill = xvar)) +
  labs(y="",x="") +
  theme_void() +
  theme(legend.position = 'none')

cplot_short=ggplot(cd[1:5,],aes(xvar,yvar)) +
  geom_col(aes(fill = xvar)) +
  labs(y="",x="") +
  theme_void() +
  theme(legend.position = 'none')
```

```{r echo=FALSE, warning=FALSE, results='hide'}
cplot1=cplot + scale_fill_uwb("quali") + labs(title='Qualitative scale "quali"')
cplot2=cplot + scale_fill_uwb("faks") + labs(title='UWB faculties "faks"')
cplot3=cplot_short + scale_fill_uwb("mono", n=5) + labs(title='Monochromatic scale  "mono"')
cplot4=cplot + scale_fill_uwb("long", n=10) + labs(title='Long scale "long"')
cplot5=cplot_short + scale_fill_uwb("bi",n=5) + labs(title='Bipolar scale "bi"')

cplots=cplot1 / cplot2 / cplot3 / cplot4 / cplot5
print(cplots)

```

-   General qualitative scale is the default.

-   Faculty scale gives the color for UWB faculties in aphabethical order

-   Monochromatic and longscales are for visualization of intensity, going from light to dark. Mono is suitable for ordinal variable with a low number of categories (up to 5) or a countinuous variable with a smaller range. Long provides a longer hue range to

-   Bi has darker poles and lighter middle values to visualise polarity (agree-disagree etc.).

The two qualitative scales (`quali` and `faks`) do not need any other arguments, just the name of the scale. The number of colors is set according to the number of categories. The ordinal scales need argment `n` which gives the number of categories. This generates the required number of steps in the color scale.

`uwb_palettes()` is used inside the `scale_fill/color_uwb()` functions and can be also called separately. For instance when creating asymmetric gradients with several levels of agreement +and a definitive disagreement.

### Option `add_na`

This option adds a grey color at the end of the scale for NA answers that are named factor levels. Real NA values are plotted as grey by default. But when we give them a name (e.g. "Don't know"), we want them to be gray instead of a color from the palette.

```{r echo=TRUE, warning=FALSE, results='hide'}
cplot1=cplot + scale_fill_uwb("quali", add_na=T) + labs(title='Qualitative scale')
cplot2=cplot + scale_fill_uwb("faks") + labs(title='UWB faculties')
cplot3=cplot_short + scale_fill_uwb("mono", n=5) + labs(title='Monochromatic scale')
cplot4=cplot + scale_fill_uwb("long", n=10) + labs(title='Long scale')
cplot5=cplot_short + scale_fill_uwb("bi",n=5) + labs(title='Bipolar scale')

cplots=cplot1 / cplot2 / cplot3 / cplot4 / cplot5
print(cplots)
```

### Option `reverse` TO BE DONE

```{r echo=TRUE, warning=FALSE, results='hide'}
```

## Examples

### Lollipop plot of a multiple choice question, labs imported from codebook

Columns `item_1` to `item_5` in the example dataset are answers to a multiple choice question where 1 indicates that respondents chose the item and 0 means not choosing it. The `prep_bat()` function takes all columns of the dataframe as items, by default. This can be changed by specifying the `vars` argument instead. The `chosen` argument specifies which value indicate choosing the item. If not provided, all values are kept (in this case, it would be 1 and NA). When `chosen` is specified, the resulting dataset can be used with `plot_bar()` or `plot_lolli()`. This example show the lollipop plot with ordered items.

```{r echo = TRUE, warning = FALSE, output = TRUE}

codebook = tibble(
  name = paste0("item_",1:5),
  title = "Which ot these statements apply to your situation?",
  subtitle = paste0("UWB students, N = ", nrow(example_data)),
  caption = "The percentages do not sum to 100 % because respondents could choose multiple answers."
  ) %>% 
  mutate(lab = case_when(name == "item_1" ~ "First statement",
                         name == "item_2" ~ "Very long second statement with automatic line break",
                         name == "item_3" ~ "Third statement",
                         name == "item_4" ~ "Fourth statement",
                         name == "item_5" ~ "Fifth statement",))

print(codebook)

 multi = prep_bat(example_data %>% select(item_1:item_5), chosen = 1)
 head(multi)
 multiplot = plot_lolli(multi)
```

## More about the color scales

### Check for color blindness

```{r echo=TRUE, warning=FALSE, results='hide'}
palette_check(c(uwb_palettes(name='quali',type = "discrete"),c_nor), plot = TRUE)
palette_check(c(uwb_palettes(name='faks',type = "discrete"),c_nor), plot = TRUE)
palette_check(c(uwb_palettes(name='mono',n=10,type = "continuous"),c_nor), plot = TRUE)
palette_check(c(uwb_palettes(name='long',n=10,type = "continuous"),c_nor), plot = TRUE)
palette_check(c(uwb_palettes(name='bi',n=9,type = "continuous"),c_nor), plot = TRUE)
```

## Resources

### theme and color palettes

https://www.r-bloggers.com/2022/06/custom-colour-palettes-for-ggplot2/ color palette generator (for scales): https://www.vis4.net/palettes/#/9\|s\|00429d,96ffea,ffffe0\|ffffe0,ff005e,93003a\|1\|1 https://projects.susielu.com/viz-palette?colors=%5B%22#1DABE6%22,%22#1C366A%22,%22#C3CED0%22,%22#E43034%22,%22#FC4E51%22,%22#AF060F%22%5D&backgroundColor=%22white%22&fontColor=%22black%22 https://rpubs.com/mclaire19/ggplot2-custom-themes

### How to create a package

### Viridis code

https://github.com/sjmgarnier/viridis/tree/master

## Notes, future development

```{r include=FALSE}
#hexagon plot
dat <- data.frame(x = rnorm(10000), y = rnorm(10000))

#ggplot(dat, aes(x = x, y = y)) {
#  geom_hex() + coord_fixed() +
#  scale_fill_viridis() + theme_bw()


#test theme
#plots to test theme
mtcars2 <- within(mtcars, {
  vs <- factor(vs, labels = c("V-shaped", "Straight"))
  am <- factor(am, labels = c("Automatic", "Manual"))
  cyl  <- factor(cyl)
  gear <- factor(gear)
})

p1 <- ggplot(mtcars2) +
  geom_point(aes(x = wt, y = mpg, colour = gear)) +
  labs(
    title = "Fuel economy declines as weight increases",
    subtitle = "(1973-74)",
    caption = "Data from the 1974 Motor Trend US magazine.",
    tag = "Figure 1",
    x = "Weight (1000 lbs)",
    y = "Fuel economy (mpg)",
    colour = "Gears"
  )
p2 <- p1 + facet_grid(vs ~ am)

#p1 + theme_uwb()
#p2 + theme_uwb()